Ages of Planets and Stars Observed by Kepler

### 2024-04-09 paper/ms.tex
update ms

To evaluate the accuracy and precision of this approach, after
applying an initial iteration of this method on the 
\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
Z.}

@@

In terms of detected planet counts, our rotation-based ages for
the Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
For the sub-gigayear bin in particular, which holds specific weight in
studies of planet evolution, while there are \nplyounggyro\ planets
with median ages below 1\,Gyr, requiring $t_{\rm
gyro}$$<$1\,Gyr at 2$\sigma$ confidence yields
\nplyounggyrotwosigma\ planets orbiting \nplhostsyounggyrotwosigma\
stars.

\subsection{The Sub-Gigayear Sample}

Given that the history of young exoplanet discovery includes a
sociological tendency to report detailed descriptions of specific
young exoplanets, we find it necessary to briefly discuss the
properties of at least the most notable members of the set of
confirmed planets that have well-measured ages.
For this exercise, we follow Table~\ref{tab:youngplanets}.

\paragraph{The Cep-Her Planets}
Kepler\,1627Ab, Kepler\,1643b, Kepler\,1974b and Kepler\,1975\,Ab are
known members of the $\approx$40\,Myr old Cep-Her association
\citep{Bouma_2022a,Bouma_2022b}.
The age of various sub-components of this association are
best measured through relative isochronal placement of the
pre-main-sequence \citep[see][]{Bouma_2022b}, and gyrochronology of
stars in these clusters show them to be $<$100\,Myr old.
The planets themselves are all on close-in orbits (5-25\,days), with
sizes from 2-4\,$R_\oplus$.
For Kepler\,1627A and Kepler\,1974, the rotation and lithium-based
ages for these stars agree within 1$\sigma$ with the cluster-based
ages.
{\bf TODO discuss Kepler-1643 old tension, ditto KOI7913...}

\paragraph{Theia~520 Planets}

Kepler\,968 and Kepler\,52 both have rotation-based ages near
350\,Myr.
They are also members of Theia~520 (CITE).
Independent work by J.~Curtis et al.~is indicating an age of XXX\,Myr for this
cluster.
This makes them the youngest multiplanet systems that have currently
been identified from the Kepler primary mission.

\paragraph{NGC~6811 Planets}
Meibom (CITE CITE) ... do we get those ones?

\paragraph{Systems with Rotation and Lithium in Agreement}
Kepler\,1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is
$\approx$100\,Myr old.

Kepler\,1928b, a 19.6\,day 1.9\,$R_\oplus$ mini-Neptune is
$\approx$140-200\,Myr old.

Kepler\,1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing
orbit is $\approx$120-180\,Myr old.

Kepler\,1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth
on a grazing orbit is $\approx$185\,Myr old.

Kepler\,1521b, a 47.2\,day 2.3\,$R_\oplus$ super-Earth is 180-210\,Myr
old.

Kepler\,1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is
230-240\,Myr old.

Kepler\,1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr
old.

Kepler\,971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is
300-370\,Myr old.

Kepler\,1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$
of its lithium age, though at half the relative uncertainty.
This system may be the youngest Earth-sized planet known, comparable to 
systems such as HD~63433d \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.

\paragraph{Other Notables}

@@

\paragraph{Sub-}

Kepler\,957b

To date, four stellar clusters in the Kepler field have yielded planet
detections.
They are:...
Young Kepler systems with the most precisely measured ages are...

(TODO: perhaps easiest in a little table.  Theia316, Theia520, CepHer,
and NGC6811).

\subsection{Ages for KOIs}

@@

@@

\startlongtable
\begin{deluxetable*}{llllrrrrr}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
	\tablecaption{Young planets and planet candidates.
		%The non-truncated machine readable versions are accessible both
		%through the online journal, and at
		%\url{https://zenodo.org/record/8327508}.
		\label{tab:youngplanets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
  -- &   -- & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
	\hline
	\input{table_subgyr_confirmed} %\\
	%\hline
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{This table includes X, Y, Z... The machine-readable version,
		available online, includes additional columns for .... 
	}
\end{deluxetable*}
%\endlongtable

>>>>

To evaluate the accuracy and precision of our equivalent measurement
technique, after applying an initial iteration on the
\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
Z.}

@@

In terms of detected planet counts, our rotation-based ages for
the Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
For the sub-gigayear bin in particular, which holds specific weight in
studies of planet evolution, while there are \nplyounggyro\ planets
with median ages below 1\,Gyr, requiring $t_{\rm
gyro}$$<$1\,Gyr at 2$\sigma$ confidence yields
\nplyounggyrotwosigma\ planets orbiting \nplhostsyounggyrotwosigma\
stars.
The stellar and planetary properties of this latter sample of young
planets are given in Table~\ref{tab:youngplanets}.

\subsection{Planets Younger Than One Billion Years}

Given that the history of young exoplanet discovery includes a
sociological tendency to report detailed descriptions of specific
young exoplanets, we find it necessary to briefly discuss the
properties of at least the most notable members of the set of
confirmed planets that have well-measured ages.
We first focus on systems reported to be in the four stellar
ensembles in the Kepler field that have yielded transiting planets:
Cep-Her, MELANGE-3, Theia~520, and NGC~6811.
We then consider the broader set of Kepler planets that are young
based on the rotational and lithium-based properties of their host
stars.

\subsubsection{Planets in Clusters}
\label{subsec:clusterplanets}

\paragraph{Cep-Her}
Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab are
known members of the $\approx$40\,Myr old Cep-Her association
\citep{Bouma_2022a,Bouma_2022b}.  The age of various sub-components of
this association are best measured through relative isochronal
placement of the pre-main-sequence \citep[see][]{Bouma_2022b}, and
gyrochronology of stars in these clusters show them to be $<$100\,Myr
old.  The planets themselves are all on close-in orbits (5-25\,days),
with sizes from 2-4\,$R_\oplus$.  For Kepler-1627A and Kepler-1974,
the rotation, lithium, and cluster-based ages all agree within
1$\sigma$.  For Kepler-1643, while the rotation and cluster ages agree
($46^{+9}_{-7}$\,Myr; \citealt{Bouma_2022b}), the lithium-based age
($198^{+98}_{-76}$\,Myr) is 2.0$\sigma$ above the nominal cluster age
for RSG-5.  For Kepler-1975Ab (KOI\,7913Ab), the lithium age may be
similarly ``old'', though with less confidence due to the large
intrinsic scatter in K dwarf lithium calibration measurements.  For
that system, the isochronal age of its binary companion provides a
more stringent and serendipitous constraint on the $\approx$40\,Myr
age.

\paragraph{MELANGE-3}
\citet{2022AJ....164...88B} described an association in the Kepler
field that was argued to be $105\pm10$\,Myr based on a combination of
rotation, lithium, and the isochronal pre-main-sequence.  Kepler-970
and Kepler-1928 harbor two transiting mini-Neptunes inside this
association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and -1928 respectively.  The former rotation-based age
being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}, where
the MELANGE-3 rotational isochrone is visually slower than
the corresponding Pleiades track.

\paragraph{Theia~520}
Kepler-968 and Kepler-52 both have rotation-based ages of
330-350\,Myr.  These stars were independently reported by
\citet{2019AJ....158..122K} to be members of an association dubbed
Theia~520 based on their spatial and kinematic correlations with
nearby stars.  While the core of this structure was previously named
UBC-1 \citep{2018A&A...618A..59C}, both Kepler planet hosts are in
more diffuse regions of the cluster reported by
\citet{2019AJ....158..122K}, so we adopt the name from the latter
study.  The nominal age of Theia~520 reported by
\citet{2019AJ....158..122K} based on isochrone fitting was
$247^{+64}_{-51}$\,Myr.  An independent analysis of the cluster's core
by \citet{2024A&A...681A..13F} using g-mode asteroseismology,
gyrochronology, and isochrone fitting found consistent results.
Ongoing spectroscopic and rotation-based studies of the broader
cluster by J.~Curtis et al.~ is broadly that the two planet-hosting
stars are indeed members of this diffuse population.  Given the broad
set of both cluster-level and individual-star evidence, this makes
Kepler-968 and Kepler-52 the youngest multiplanet systems that have
currently been identified from the Kepler primary mission.
We leave detailed discussion of the planet's properties to the latter 
group.

\paragraph{NGC~6811}
\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
We report rotation-based ages for these systems of
$1419^{+657}_{-353}$\,Myr and $873.75\pm{112}$\,Myr respectively.
For Kepler-66, $T_{\rm eff}$=5898\,K, and $P_{\rm
rot}$=10.25$\pm$0.10\,days.
{\it LGB TODO: check again after dereddening...}

\subsubsection{Systems with Rotation and Lithium in Agreement}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age greatly
increases, due to the independence of the two measurement techniques.
The empirical precision for the inferred ages as a function of $P_{\rm
rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
respectively.  The region of greatest overlap is dictated by the
timescale for lithium depletion: for stars with temperatures between
4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
independent two-sided constraints.

The youth of most of these systems, to our knowledge, is being
well-constrained for the first time in this work.  Here we limit
discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
mentioned above.

The new young planets
include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
(Kepler-1529b $\approx$100\,Myr,
-1521b $\approx$195\,Myr,
-1467b $\approx$235\,Myr,
-955b $\approx$400\,Myr, and
-523b $\approx$440\,Myr),
a single Neptune-sized planet
(Kepler-971b $\approx$335\,Myr),
five 1.1-1.8\,$R_\oplus$ super-Earths
(Kepler-1930b $\approx$150\,Myr,
-1313b $\approx$185\,Myr,
-1565b $\approx$205\,Myr,
-1532b $\approx$400\,Myr,
-1481b $\approx$400\,Myr),
and three Earth or sub-Earth-sized planets
(Kepler-1933b $\approx$300\,Myr,
-1561b $\approx$420\,Myr,
-1482b $\approx$490\,Myr).
All of these planets are on orbital periods below 50\,days.

Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
though at half the relative uncertainty.  Based on these lines of
evidence, this system may be the youngest Earth-sized planet known,
comparable to systems such as HD~63433d \citep[1.1\,$R_\oplus$,
$414\pm23$\,Myr][]{2024AJ....167...54C}.

A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.

%  Kepler-1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is $\approx$100\,Myr old.
%  Kepler-1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$120-180\,Myr old.
%  Kepler-1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$185\,Myr old.
%  Kepler-1521b, a 47.2\,day 2.3\,$R_\oplus$ mini-Neptune is 180-210\,Myr old.
%  Kepler-1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is 230-240\,Myr old.
%  Kepler-1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr old.
%  Kepler-971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is 300-370\,Myr old.
%  
%  Kepler-1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
%  rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of
%  its lithium age, though at half the relative uncertainty.  Based on
%  these lines of evidence, this system may be the youngest Earth-sized
%  planet known, comparable to systems such as HD~63433d
%  \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.
%  
%  Kepler-1532b is a 1.1\,day 1.3\,$R_\oplus$ super-Earth.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
%  Kepler-955b is a 14.5\,day 2.7\,$R_\oplus$ mini-Neptune.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
%  Kepler-1481b, a 5.9\,day 1.1\,$R_\oplus$ super-Earth has a
%  rotation-based age of $\approx$400\,Myr, and a $2\sigma$ lithium
%  detection that, upon visual inspection of the spectrum, is
%  believeable.  The implied lithium age is marginally consistent with
%  the rotation-based age, although perhaps with some tension.
%  Kepler-447b is supposedly a 32\,$R_\oplus$ planet (grazing) on a
%  7.8\,day grazing orbit.  Rotation and lithium independently find a
%  420-440\,Myr age.
%  Kepler-1561b is a 0.9\,$R_\oplus$ sub-Earth, on a 1.0\,day orbit.
%  Rotation points to a 420\,Myr age.  The detected EW$_{\rm Li}$ is
%  consistent, although given the star's temperature, not particularly
%  constraining.
%  The same is true for the 1.9\,$R_\oplus$ and 5.8\,day super-Earth Kepler-523b, and the 
%  1.0\,$R_\oplus$, 12.2\,day Earth-sized Kepler-1482b.

\subsubsection{Systems with Rotation and Lithium in Disagreement}

For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
rotational broadening of the spectrum ($v\sin
i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
rotation period.  However, if this star were $\lesssim$500\,Myr old,
lithium should have been detectable at $>$50\,m\AA.  Lithium is
marginally if at all present in the spectrum ({\bf  EW$_{\rm
Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
\citep{Morton_2016}, inspecting the data validation reports this is
somewhat surprising.  The light curve shows greater scatter in-transit
than out of transit, appears near-grazing, and has a suggestion of a
strong odd-even depth difference.  This is a surprising system that
merits further attention.

\subsubsection{Need spectra}
Kepler-1938...
Kepler-762... (a HJ!)
Kepler-1320

\subsubsection{Other Notables}

@@

@@

% 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
Our rotation analysis blindly recovered the 14 known Kepler planets in
clusters (Section~\ref{subsec:clusterplanets}), and provided accurate
independent ages for their host stars.

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrr}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
	\tablecaption{Young planets and planet candidates.
		%The non-truncated machine readable versions are accessible both
		%through the online journal, and at
		%\url{https://zenodo.org/record/8327508}.
		\label{tab:youngplanets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
	\hline
	\input{table_subgyr_confirmed} %\\
	%\hline
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{This table includes X, Y, Z... The machine-readable version,
		available online, includes additional columns for .... 
	}
\end{deluxetable*}
%\endlongtable
